A Journey from Wild to Textbook Data to Reproducibly Refresh the Wages Data from the National Longitudinal Survey of Youth Database

### 2022-02-17 paper/paper.Rmd
replaced rlm plot with same set of individuals as fig 2 - looks better

In the next step, we explore the mean hourly wage data. In this case, we only explore the wages data of respondents that have the highest completed grade of up to $12^{th}$ grade.  We employ visualization techniques to perform the IDA as described next.

```{r sample-plot, echo = FALSE, fig.cap="Longitudinal profiles of wages for a random sample of 36 individuals in the pre-cleaned data. There is considerable variation in wages. Some individuals  (2799, 11041, 11146) are only measured for a short period. Some individuals (8296, 9962) possibly have errors in wages in some years.", fig.height=8, fig.width=10, out.width="100%"}
```

@@

We randomly take 20 samples from the data and plot them, as shown in Figure \ref{fig:sample-plot}.<!--AE comment: "It" implies... Please specify what "it" refers to. This could be done by adding onto the first sentence... "as shown in Figure 2, which shows that some respondents have high variability in wages, particularly 5, 7, and 11." However, 7 and 11 don't seem to have particularly high variability, especially when compared to 18. This whole paragraph needs to be fixed/updated.--> It shows that these respondents have a lot of variability in wages, for example, the IDs in panel numbers 5, 7, and 11. The plot also implies that the samples have a different pattern of mean hourly wages. Some have had flat wages for years but had a sudden increase in one particular year, then it gone down again, while the others experienced an upsurge in their wage, for instance, <!--AE comment: Not panel 9 --> the IDs in panel 9. 
<!--I think there is a narrative step missing here, which is that looking at those samples led you to look at plots that show values for the wage variable over all participants and all years. The phrase “the summary plots” was not enough description for me to know what to expect when I flipped to Figure 3. Please be explicit about what the three plots were, why they were made, and what you were looking for. I think there is another narrative step missing about Figure 3C, which is that after you saw there were outliers you tried to identify which respondents had the extreme values. Again, please spell this out in the text. When you return to the plots for the entire dataset in Figure 5, please make parallel comments. -->
However, when checking the summary plots (Figure \ref{fig:feature-plot}), we found that some observations had exceptionally high wages, the maximum value of the wages unbelievably reached about $60,000/hour. Besides, some of them, for example respondents in Figure \ref {fig:feature-plot} C, had experienced unusual wages only in certain year.

@@

Figure \ref{fig:flow-chart-blind} summarizes the steps taken to go from raw to input to valid data [@LooMarkvander2018Sdcw] to create a refreshed wages data set.

>>>>

```{r sample-plot, echo = FALSE, fig.cap="Longitudinal profiles of wages for a random sample of 36 individuals in the pre-cleaned data. There is considerable variation in wages. Some individuals  (2799, 11041, 11146) are only measured for a short period. Some individuals (8296, 9962) possibly have errors in wages in some years, because of the extreme fluctuation.", fig.height=8, fig.width=10, out.width="100%"}
```

@@

In the next step, we explore the mean hourly wage data of samples of individuals. The purpose is to examine the common patterns and check the quality. A random sample of 36 individuals is chosen (using the `sample_n_keys` function in `brolgar`). Their longitudinal profiles are plotted, faceted by `id` and using free $y$ scales, so that the individual patterns can be examined (Figure \ref{fig:sample-plot}).<!--AE comment: "It" implies... Please specify what "it" refers to. This could be done by adding onto the first sentence... "as shown in Figure 2, which shows that some respondents have high variability in wages, particularly 5, 7, and 11." However, 7 and 11 don't seem to have particularly high variability, especially when compared to 18. This whole paragraph needs to be fixed/updated.--> There is a lot of variability from one individual to another, and substantial fluctuation in wages at different times for most individuals. Some individuals  (2799, 11041, 11146) are only measured for a short period. Some individuals (8296, 9962) possibly have errors in wages in some years, because of the extreme fluctuation. These need to be inspected more closely.

 <!--AE comment: Not panel 9 --> the IDs in panel 9. 
<!--I think there is a narrative step missing here, which is that looking at those samples led you to look at plots that show values for the wage variable over all participants and all years. The phrase “the summary plots” was not enough description for me to know what to expect when I flipped to Figure 3. Please be explicit about what the three plots were, why they were made, and what you were looking for. I think there is another narrative step missing about Figure 3C, which is that after you saw there were outliers you tried to identify which respondents had the extreme values. Again, please spell this out in the text. When you return to the plots for the entire dataset in Figure 5, please make parallel comments. -->

Figure \ref{fig:feature-plot}s shows an alternative way to check the data quality. Plot (A) is the spagetti plot where all profiles are shown, and it can be seen that there are unbelievably high wage values (up to $60,000/hour) for some individuals mostly around 1990. Plot (B) shows side-by-side boxplots of the three number summaries (minimum, median and maximum) for all individuals. This tells us that there are a number of individuals with unbelievably high maximum wages. Plot (C) shows the profile for an individual, with not such a high maximum wage but still indicates a problem: their wages are consistently low except for one year where they earned close to $1200/hour. This does not seem to be reasonable, and leads us to use a procedure to detect and fix these temporal anomalies.

@@

There are many steps and decisions made to go from raw to input to valid data. Figure \ref{fig:flow-chart-blind} summarizes these in order to create a refreshed wages data set.

### 2022-02-17 paper/paper.tex
replaced rlm plot with same set of individuals as fig 2 - looks better

In the next step, we explore the mean hourly wage data. In this case, we only explore the wages data of respondents that have the highest completed grade of up to \(12^{th}\) grade. We employ visualization techniques to perform the IDA as described next.

@@

\caption{Longitudinal profiles of wages for a random sample of 36 individuals in the pre-cleaned data. There is considerable variation in wages. Some individuals  (2799, 11041, 11146) are only measured for a short period. Some individuals (8296, 9962) possibly have errors in wages in some years.}\label{fig:sample-plot}
\end{figure}

We randomly take 20 samples from the data and plot them, as shown in Figure \ref{fig:sample-plot}. It shows that these respondents have a lot of variability in wages, for example, the IDs in panel numbers 5, 7, and 11. The plot also implies that the samples have a different pattern of mean hourly wages. Some have had flat wages for years but had a sudden increase in one particular year, then it gone down again, while the others experienced an upsurge in their wage, for instance, the IDs in panel 9.
However, when checking the summary plots (Figure \ref{fig:feature-plot}), we found that some observations had exceptionally high wages, the maximum value of the wages unbelievably reached about \$60,000/hour. Besides, some of them, for example respondents in Figure \ref {fig:feature-plot} C, had experienced unusual wages only in certain year.

@@

Figure \ref{fig:flow-chart-blind} summarizes the steps taken to go from raw to input to valid data (M. van der Loo and de Jonge 2018) to create a refreshed wages data set.

>>>>

@@

\caption{Longitudinal profiles of wages for a random sample of 36 individuals in the pre-cleaned data. There is considerable variation in wages. Some individuals  (2799, 11041, 11146) are only measured for a short period. Some individuals (8296, 9962) possibly have errors in wages in some years, because of the extreme fluctuation.}\label{fig:sample-plot}
\end{figure}

In the next step, we explore the mean hourly wage data of samples of individuals. The purpose is to examine the common patterns and check the quality. A random sample of 36 individuals is chosen (using the \texttt{sample\_n\_keys} function in \texttt{brolgar}). Their longitudinal profiles are plotted, faceted by \texttt{id} and using free \(y\) scales, so that the individual patterns can be examined (Figure \ref{fig:sample-plot}). There is a lot of variability from one individual to another, and substantial fluctuation in wages at different times for most individuals. Some individuals (2799, 11041, 11146) are only measured for a short period. Some individuals (8296, 9962) possibly have errors in wages in some years, because of the extreme fluctuation. These need to be inspected more closely.

the IDs in panel 9.

Figure \ref{fig:feature-plot}s shows an alternative way to check the data quality. Plot (A) is the spagetti plot where all profiles are shown, and it can be seen that there are unbelievably high wage values (up to \$60,000/hour) for some individuals mostly around 1990. Plot (B) shows side-by-side boxplots of the three number summaries (minimum, median and maximum) for all individuals. This tells us that there are a number of individuals with unbelievably high maximum wages. Plot (C) shows the profile for an individual, with not such a high maximum wage but still indicates a problem: their wages are consistently low except for one year where they earned close to \$1200/hour. This does not seem to be reasonable, and leads us to use a procedure to detect and fix these temporal anomalies.

@@

There are many steps and decisions made to go from raw to input to valid data. Figure \ref{fig:flow-chart-blind} summarizes these in order to create a refreshed wages data set.
